Sensitivity Analysis

### 2024-05-15 SFE_article_Sensitivity_Analysis_short - Chamomile/Article_Sensitivity_Analsysis_short-Chamomile.tex
15.05

The sensitivity analysis is a tool to understand the parametric dependence in the dynamic behaviour of the analysed system. The presented formulation involves derivative-based local sensitivity analysis of the model solution with respect to selected parameters and controls. The sensitivity equations can be obtained in various ways, but the automatic differentiation technique was implemented to obtain the sensitivity equations in this work. By identifying which parameters are influential at each period of the simulation, local sensitivity analysis can bring valuable information about process modelling, design of experiments or model reduction. The local sensitivity analysis techniques consider only a small region of parameter space, and the conclusions derived from such an analysis are limited to local conditions unless the discussed system is a linear model.

In this study, the local sensitivity analysis method was introduced to explore the dynamics of the supercritical extraction system, which consists of a set of partial differential equations. The sensitivity analysis was used to evaluate the influence of pressure increase on the state-space consisting of the solutes concentration in solid and liquid phases, the fluid's enthalpy density, system pressure, and extraction yield. Every change in the control variables leads to a state-space change, eventually affecting the process performance.

At given operating conditions, the step change of pressure enhance the mass transfer, which leads to observation faster loss of solute by the solid particles and consequently to negative sensitivities. Analogous response is visible in the fluid phase, where the concentration of solute increase, which is visible as positive sensitivities. As more solute is transported to the fluid phase in shorter time, the extraction yield is also improved.

@@

This information can be utilized to identify which control variables influence the extraction yield the most. The controls with high sensitivities on the extraction can be used to investigate to find optimal operating conditions from an economic point of view.

>>>>

Sensitivity analysis is a tool to understand the parametric dependence on the dynamic behaviour of the analysed system. The presented formulation involves derivative-based local sensitivity analysis of the model solution with respect to selected parameters and controls. The local sensitivity analysis techniques consider only a small region of parameter space, and the conclusions derived from such an analysis are limited to local conditions unless the discussed system is a linear model. The sensitivity equations can be obtained in various ways, but the automatic differentiation technique was implemented to obtain the sensitivity equations in this work. Local sensitivity analysis can bring valuable information about process modelling, design of experiments or model reduction by identifying which parameters are influential and how these influence changes in time.

This study used sensitivity analysis to evaluate the influence of pressure increase on the state-space, which consisted of the concentration of the solute in solid and liquid phases, the fluid's enthalpy density, system pressure, and extraction yield. At given operating conditions ($35^\circ C$, 150 bar and 5 $\cdot 10^{-5}$ kg/s), the step change of pressure enhance the mass transfer, which leads to faster loss of solute from solid particles and consequently to negative sensitivities. An analogous response is visible in the fluid phase, where the concentration of solute increases and is visible as positive sensitivities. As more solute is transported to the fluid phase, the extraction yield is improved and characterised by positive sensitivities.

This information can be utilised to identify which control variables influence the extraction yield the most. The controls with high sensitivities to the extraction can be used to investigate optimal operating conditions from an economic point of view.

@@

### 2024-05-15 SFE_article_Sensitivity_Analysis_short - Chamomile/Sections/Sensitivity_Pressure.tex
14.05
15.05

	As discussed in Chapter \ref{CH:Governing_equations_chapter}, a small pressure wave propagates at the speed of sound relative to the flow. If the flow velocity is relatively low, all pressure changes are hydrodynamic (resulting from velocity motion) rather than thermodynamic. The Low Mach-number assumption leads to instant propagation of the thermodynamic pressure throughout the system. A single pressure value can be considered for the entire system, as all changes occur simultaneously throughout the machine. Figure \ref{fig:Sensitivty_P_P} illustrates a step function representing the pressure change.

@@

@@

	The pressure change influences the temperature of the fluid in the middle of the computational domain, but the values on boundaries can be defined in multiple ways. Applying the Dirichlet boundary conditions assumes a fixed temperature value at the inlet, which leads to a possible thermal gradient propagating along the system. In contrast, Neumann boundary conditions would dictate that the heat flux at the boundaries be held constant. In this work, the Neumann boundary conditions have been applied, which leads to a uniform response and ensures that the temperatures at the inlet and in the middle of the extractor are the same.

@@

	Figure \ref{fig:Sensitivty_P_CS} shows the sensitivity of the solute concentration in the solid phase with respect to changes in pressure throughout a fixed bed in a supercritical fluid extraction process. As discussed in Chapter \ref{CH: Continuity}, the velocity of a fluid is inversely related to its density, which suggests that with higher fluid density, the velocity decreases. This leads to an extended residence time, ergo, a longer interaction between the solute and the solvent.

	Initially, the extraction process is in the kinetically-controlled regime, where the concentration gradient is high, and the limiting factor is the solute solubility. As explained in ({\color{red}article 1}), the system is considered to be far from the saturation, which can explain low system response at the beginning of the process.

	When the concentration gradient diminishes, and the extraction moves from the kinetic-controlled to the diffusion-controlled regime, the system response becomes more profound. ({\color{red}article 1}) obtained a linear correlation for $D_i^R$, which is incorporated in this model. $D_i^R$ increases together with fluid density, which explains negative sensitivities. The negative sign can be interpreted as a faster loss of a solute from a solid phase, which corresponds to enhanced mass transfer.

	Over time, the amount of solute becomes a limiting factor and the pressure change has a lower effect on the system and the sensitivities increase. Eventually, sensitivities asymptotically approach zero. The solute concentration in the solid phase has been reduced, and further changes in pressure do not influence the remaining concentration. This behaviour reflects that in the later extraction phase, the kinetics of extraction are controlled by the internal diffusion of the solute from solid particles rather than the conditions of the supercritical fluid.

@@

	Figure \ref{fig:Sensitivty_P_CF} shows the sensitivity of the solute concentration in the fluid phase with respect to the pressure change. As explained above, an increase in pressure enhances the mass transfer, resulting in the concentration front moving through the system. The system response is initially low, reflecting the idle period discussed above. Later, the sensitivities increased due to enhanced mass transfer. As the fluid phase is mobile, the advection affects the sensitivities, which causes the sensitivities to move along the system analogously to the solute in the fluid phase.

	After the growth phase, the sensitivities start to decrease, which can be explained by the decrease in the concentration gradient. As seen in Figure \ref{fig:Sensitivty_P_CF}, sensitivities decline towards zero. This asymptotic trend suggests the driving force for the mass transfer—the concentration gradient—has been minimized, and changes in pressure no longer significantly impact the solute concentration in the fluid phase.

@@

	Figure \ref{fig:Sensitivty_P_y} illustrates how sensitive the extraction yield is to pressure change. Initially, the sensitivity curve stays almost flat, suggesting a latency in the system's response to pressure changes. A minor decrease in sensitivity arises due to a decrease in the velocity of the fluid phase. As the fluid moves slower, it reaches the end of the extractor later, which causes negative sensitivities.

	As the process continues and the solute reaches the extractor's outlet, the sensitivity curve increases rapidly, indicating an improvement in the process efficiency. This increase in sensitivity can be associated with the enhanced solubility and mass transfer imparted by higher pressures. The peak in $\frac{dy}{dP}$ reflects when the deviation from the original system is the largest.

	Beyond the peak, the sensitivity declines, converging towards zero. This descending curve represents a depletion phase: as the available solute in the solid phase diminishes, the potential for further increasing the yield through pressure alone is reduced. The remaining concentration gradient becomes a limiting factor, and the enhanced mass transfer no longer plays a dominant role compared to the original system.

>>>>

	As discussed in Chapter \ref{CH:Governing_equations_chapter}, a small pressure wave propagates at the speed of sound relative to the flow. If the flow velocity is relatively low, all pressure changes are hydrodynamic (resulting from velocity motion) rather than thermodynamic. The Low Mach-number assumption leads to instant propagation of the thermodynamic pressure throughout the system. A single pressure value can be considered for the entire system, as all changes occur simultaneously throughout the device. Figure \ref{fig:Sensitivty_P_P} illustrates a step function representing the pressure change.

@@

	\todo{caption and label}

@@

	The pressure change affects the fluid's temperature inside the computational domain, but boundary conditions define the values on boundaries. Applying the Dirichlet boundary conditions assumes a fixed temperature value at the inlet, which can lead to a thermal gradient propagating along the system. In contrast, Neumann boundary conditions would dictate the heat flux at the boundaries. In this work, the Neumann boundary conditions equal to zero have been applied, which leads to a uniform response and ensures that the temperatures at the inlet, the outlet, and the middle of the extractor are the same.

@@

	Figure \ref{fig:Sensitivty_P_CS} shows the sensitivity of the solute concentration in the solid phase with respect to pressure change along a fixed bed in a supercritical fluid extraction process. As discussed in Chapter \ref{CH: Continuity}, the velocity of a fluid is inversely proportional to its density, which suggests that with higher fluid density, the velocity decreases. This leads to an extended residence time, ergo, a longer interaction between the solute and the solvent. Initially, the extraction process is in the kinetically-controlled regime, where the concentration gradient is high, and the limiting factor is the solute solubility. As discussed in ({\color{red}article 1}), the system is considered to be far from saturation, which can explain the low system response at the beginning of the process. The system response becomes more evident when the concentration gradient diminishes, and the extraction moves from the kinetic-controlled to the diffusion-controlled regime. The negative sign can be interpreted as a faster loss of a solute from the solid phase, which corresponds to enhanced mass transfer. Over time, the amount of solute becomes a limiting factor, and the pressure change has a lower effect on the system, increasing the negative sensitivities. Eventually, sensitivities approach zero asymptotically. The solute concentration in the solid phase has been reduced, and further changes in pressure have a low influence on the state space.

@@

	Figure \ref{fig:Sensitivty_P_CF} shows the sensitivity of the solute concentration in the fluid phase with respect to the pressure change. As analysed above, an increase in pressure enhances the mass transfer, resulting in the concentration front moving through the system. The system response is initially low, reflecting the idle period discussed above. Later, the sensitivities start to increase due to faster solute loss from the solid phase, which indicates that more solute is transported to the fluid phase. When the amount of solute in the solid phase becomes a limiting factor, the extraction rate slows down, sensitivities decline asymptotic towards zero, and the impact of the pressure change gradually decreases. As the fluid phase is mobile, the advection affects the sensitivities, which causes the sensitivities to move along the system analogously to the solute in the fluid phase.

@@

	Figure \ref{fig:Sensitivty_P_y} illustrates how sensitive the extraction yield is to the pressure change. Initially, the sensitivity curve stays almost flat, suggesting a latency in the system's response to pressure changes. Due to the decreased velocity of the fluid, the solute reaches the extractor's outlet later, which causes minor negative sensitivities to appear. The process continues, and the sensitivity curve increases rapidly when the solute reaches the extractor's outlet. The positive yield sensitivity indicates an improvement in the process efficiency, which is directly related to enhanced mass transfer. The peak in $\frac{dy}{dP}$ reflects when the deviation from the original system is the largest. Beyond the peak, the sensitivity declines and converges towards zero. The concentration gradient becomes a limiting factor, and the enhanced mass transfer no longer plays a dominant role compared to the original system.
